---
title: "Homework 1 - answers"
author: "Almog Angel"
date: "01/12/2024"
output: html_document
---

In the following tasks you will use statistics and sequence alignment to analyze real COVID-19 world data.

- Load packages
```{r}
library(tidyverse)

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Biostrings")

library(Biostrings)

BiocManager::install("msa")

library(msa)
library(msaR)
library(seqinr)
library(ape)
```

- Set your working directory
```{r}
# Enter your code here:
path <- getwd()
setwd(path)
```

## Task 1 - Data cleaning

- Read "COVID19_line_list_data.csv" into a variable called "covid_data"
```{r}
# Enter your code here:
covid_data <- read.csv("COVID19_line_list_data.csv")
```

- Take a look at the data
```{r}
View(covid_data)
```

- Find a way to remove empty columns:
```{r}
# Enter your code here:
covid_data %>% 
          discard(~all(is.na(.) | . =="")) -> covid_data

```

- Find a way to deal with missing values (NAs) in  "age" and "gender" columns
```{r}
# Enter your code here:

```

- Assuming 'death' is a binary variable (0 for no, 1 for yes) that represent categorical data
- Convert 'death' to a factor class that contains either "yes" or "no"
```{r}
# Enter your code here:

```


## Task 2 - Descriptive statistics analysis 

- Print a descriptive summary of "age"
```{r}
# Enter your code here:

```

- Distribution:
- (1) Generate histogram for "age" distribution and color the bars by "death"
* Use: position = 'identity', bins = 50, alpha = 0.5
```{r}
# Enter your code here:

```

- (2) Generate bar plot that visualize the number of dead and alive per gender
```{r}
# Enter your code here:

```


## Task 3 - Statistical Testing


- Run Chi-squared test to investigate the association between gender and death
```{r}
# Enter your code here:

```

- Why do we use Chi-squared and not t-test?
```{}
Type your answer here:

```

- Define the Null Hypothesis of the Chi-squared Test in this example
```{}
Type your answer here:

```

- What is the p-value? Is it significant (alpha = 0.05)? Write your conclusions from the statistical test results.
```{}
Type your answer here:

```


According to the media, older people are more likely to die from COVID-19.
- Test this claim statistically
- Frame your null and alternative hypothesis here:
```{}
My null hypothesis:

My alternative hypothesis:

```

- Check if the assumption of normality is valid for your data (use alpha of 0.05)
```{r}
# Enter your code here:

```

- Run a statistical test:
- If you assume normality -> use t-test 
- else -> use Wilcoxon rank-sum test 
```{r}
# Enter your code here:

```

- Report the p-value. Is it significant? Write your conclusions from the statistical test results.
```{}
Type your answer here:

```

- Read "owid-covid-data.csv" from "https://covid.ourworldindata.org/data/owid-covid-data.csv" into a variable called "covid_world"
```{r}
# Enter your code here:

```

- Take a look at the data
```{r}
View(covid_world)
```


Your friend suggest that countries with high Human Development Index (HDI) have a greater risk to die from COVID-19

- Make a second table and call it "covid_hdi_vs_deaths":
  (1) Select rows with "date" equal to "2024-01-21"
  (2) Select the following columns: "location", "human_development_index", "population", "total_deaths"
  (3) Remove rows with NAs
  (4) Calculate the total deaths per million people to a column named "total_deaths_per_million" (round the number)
```{r}
# Enter your code here:

```

- Use a scatterplot to visualize the relations between HDI and the total deaths per million
```{r}
# Enter your code here:

```

- Calculate the Pearson and Spearman correlation between HDI and the total deaths per million
```{r}
# Enter your code here:

```

- Report the Spearman correlation coefficient, the strength of the correlation (weak/moderate/strong/very strong) and the direction (negative/positive)
```{}
Type your answer here:

```

- Do you agree with the claim that increased in HDI causes high COVID-19 mortality? (write in detail)
```{}
Type your answer here:

```

- Find another feature in "covid_world" that explain the correlation between HDI and COVID-19 mortality
* Hint: you already have the answer in the first part of the task
```{r}
# Enter your code here:

```

```{}
Write your explanation here: 

```


## Task 4 - Sequence Alignment

- Take a look at the "Biostrings" package vignettes
```{r}
browseVignettes("Biostrings")
```

- Read the file "covid_spike_variants.fasta"
This file contain the amino acids sequence of the COVID-19 spike protein from different variants

- Read the file using the correct function from the package "Biostrings" and assign to a variable called "variants"
```{r}
# Enter your code here:

```

- How many amino acids are in the Alpha variant?
```{}
Type your answer here:

```

- Read the documentation for the Multiple Sequence Alignment (msa) function from the package "msa"
```{r}
?msa()
```

- Run MSA for the sequences in "variants" and assign the results to "variants.msa"
```{r}
# Enter your code here:

```

- Take a look at the results using the package "msaR" (notice: you can scroll right and left to see all the sequence)
```{r}
msaR(DNAMultipleAlignment(all.spikes.msa), colorscheme = "taylor")
```

- (1) Which amino-acids appear in position #13? (type "AMINO_ACID_CODE" in the console)
- (2) Write an example of a SNP that can cause the change in amino acid as we see in the Epsilon variant
```{}
Type your answer here:

```

Phylogenetic tree for the COVID19 variants:
- Use the package "seqinr" to generate a distance matrix from the MSA results and save it in a variable called "distMat"
```{r}
# Enter your code here:

```

- Use the package "ape" to apply "neighbor-joining" (nj) clustering algorithm to construct a phylogenetic tree
- Save the results in a variable called "ptree"
```{r}
# Enter your code here:

```

- Look at the phylogenetic tree and answer the following questions:
```{r}
ape::plot.phylo(ptree)
```

- (1) Which variant is the closest to Gamma?
- (2) Which pair seems to emerged from the Alpha variant?
      (a) Epsilon and Delta
      (b) Beta and Gamma
      (c) Epsilon and Delta
      
```{}
Type your answer here:

```

- Save this R Markdown as HTML before submitting
(1) Go to "Knit" > "Knit to HTML"
(2) Change the file name to your IDs (i.e, "123456789_123456789.html")